---
title: 'CLIMARK Tutorial: Get forecast for region'
author: "Victoria Scholl"
date: "6/20/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      results = 'hide')
```

# Introduction

This tutorial pulls weather data from the aWhere API for a specific timespan and region. 

# Code 

## Install / load R packages

First, install the aWhere R packages. 

```{r install_aWhere_packages, eval=FALSE}
# install aWhere R packages
library(devtools)
devtools::install_github("aWhereAPI/aWhere-R-Library")
devtools::install_github("aWhereAPI/aWhere-R-Charts")
```

Load the R packages that contain functions used in this script.

```{r}
library(dplyr)
library(ggmap)
library(tibble)
library(wicket)
library(ggplot2)
library(aWhereAPI)
library(aWhereCharts)
```

## Define input files and parameters

Define your working directory. This is the place on your computer where your input files are located for this tutorial, and it is also where the output files will be written. Instead of typing out the entire path to every file that we need ("absolute" paths), you can just use their filenames or locations within this working directory ("relative" paths) since we are setting this working directory as our starting point for the rest of the tutorial. 

```{r}
# working directory - where input files are located and outputs will be saved.
working.dir <- "~/Documents/aWhere/"

# set the working directory 
setwd(working.dir) 

# set the working directory for R Markdown too
knitr::opts_knit$set(root.dir = working.dir)
```

Use the *source* function to load R functions that are located on your local machine. For this tutorial, there are a few functions within the *0-supporting_functions.R* file. There is a zero in the beginning of this filename so it appears at the top when files are sorted alphabetically. 

```{r load.functions}
# load external R functions in local file
source(paste0(working.dir, "0-supporting_functions.R"))
```

To pull data from the aWhere API, you'll need a Consumer Key and Secret, like a username and password. Put these in a text file with the following format: line 1 is the Consumer Key, line 2 is the Consumer Secret, and line 3 is a blank line. Set the *credentials.file* variable equal to the name of this text file. Place your credentials file in your working directory. 

```{r set.credentials.file}
# filename containing your aWhere credientials (key and secret)
credentials.file <- "credentials.txt"

# load the aWhere API credentials file 
aWhereAPI::load_credentials(paste0(working.dir, credentials.file))
```

What are the starting and ending days that define the duration of the forecast? Also specify the range of years over which to calculate the long-term normal values. 

```{r}
# starting and ending days must have this format: "YYYY-MM-DD"
day.start <- as.character(Sys.Date()) # today
day.start <- "2018-06-19" # specific date

day.end <- "2018-06-24" # specific date
day.end <- as.character(Sys.Date() + 7) # today plus 7
day.end <- as.character(as.Date(day.start) + 7) # start date plus 7

# combine the days into a single vector 
days <- c(day.start, day.end)


# "years"" is a vector with the starting year in position 1 and the ending
# year in position 2, for the long-term analysis. 
years <- c(2010, 2017)
```

This script not only pulls forecast data on a daily timestep; it also calculates aggregated forecast summaries. For a 3-day forecast, the forecasted daily precip will be summed over the course of three days. For a 7-day forecast, the forecasted daily precip will be summed over the course of seven days. Here, specify which n-day forecasts you would like to aggregate. 

```{r}
# define the length of forecast(s) to generate in a vector. 
# For a 7-day and 3-day forecast, n.day.forecasts <- c(7, 3)
n.day.forecasts <- c(7, 3)
```

Specify the template data filename containing the geographic location data.

```{r}
# define template data filename
template.file <- "CLIMARKonlyWardTemplate.csv"
``` 

To select subarea(s) of interest, list their names in this vector. For now, these subareas are limited to ward names. To generate a forecast for the entire region instead, set *subarea.select* to "ENTIRE_REGION". 

```{r}
subarea.select <- "ENTIRE_REGION"
subarea.select <- c("LOIYANGALANI", "GOLBO")
subarea.select <- "KARARE" 
```

Specify the base filename for outputs. It currently incorporates the name(s) of the subarea(s) of interest, but you can set it to be anything. 

```{r}
filename.out <- paste("AOI_Forecast",
                      paste(subarea.select, collapse="_"),
                      sep = "_")
```

When plotting the weather data, we will first display a geographic base map. Define the latitude, longitude, and zoom values for the base map. For the entire CLIMARK region, a latitude of 2.5, longitude of 38, and zoom of 7 are appropriate values. 

```{r}
# base map location and zoom values for mapping the forecast data. 
map.lat <- 2.5
map.lon <- 38
map.zoom <- 7
```

Set thresholds for different variables during mapping. For instance, we want our mapped range of precipitation data to be from 0 to 300 mm. 

```{r}
thresh.precip.max <- 300 # [mm]
thresh.precip.min <- 0 # [mm]
```

## Processing steps 

Read the template data. Filter the data to keep only the grid cells within the subarea of interest. 

```{r}
# read the template data 
template.place <- utils::read.csv(paste0(working.dir, template.file)) 

# filter the template for subarea(s) of interest
if (!identical(subarea.select, "ENTIRE_REGION")){ 
  
  template.place <- template.place %>% 
    dplyr::filter(WARD_NAME %in% subarea.select) 
  
} 
```

Pull the aWhere weather data for the specified location and timespan. The R function that we use to pull the data is *generateaWhereDataset*, which comes from the *aWhereCharts* R package. 

```{r pull.weather.data}
# query the aWhere API to get forecast data and write to file 
forecast.all <- GetForecastData(template.place, days, years, 
                                write.file = TRUE, filename.out = filename.out)
```

\pagebreak

Let's take a look at the first 10 rows of forecast data.
For a single grid (identified by *locationid*), there are seven rows of forecast data (one per day, identified by *date*). The forecasted precipitation amount (mm) for the grid cell each day is in the *precipitation.amount* column, while the long-term average precipitation amount is in the *precipitation.average* column. 

```{r results=TRUE}
head(forecast.all %>% 
       dplyr::select(locationid, date, precipitation.amount, 
                     precipitation.average), n = 10)
```

For each grid cell, let's add up the forecasted precipitation amount for an n-day forecasted precipitation. Call the *GetForecastSummary* function from the "0-supporting_functions.R" file to aggregate these daily precipitation forecast data into n-day summaries. 

```{r}
# calculate the n-day forecast summaries,
# aggregate weather data over the specified number of days.
forecasts.n <- GetForecastSummary(forecast.all, n.day.forecasts, 
                                  template.place)
```

Let's take a look at the aggregated n-day forecast summaries: 

```{r results=TRUE}
head(forecasts.n %>%
       dplyr::select(locationid, n.day, precip.amount.sum, 
                     precip.avg.sum))
```


# Map forecast summaries

To visualize the forecasted weather data geographically, let's plot the data on a map.

```{r results=TRUE}
# create the base map using the parameters defined earlier
base.map = get_map(location = c(lon = map.lon, 
                                lat = map.lat), 
                   zoom = map.zoom, 
                   color = "bw")

# display map of region
gg.map <- ggmap(base.map)
#gg.map
```

```{r}
# create a data frame for the variable thresholds.
# access the variable by referencing the column (thresholds$precip)
# the minimum values is in position 1, maximum value in position 2
thresholds <- as.data.frame(c(thresh.precip.min, thresh.precip.max))
colnames(thresholds) <- "precip"
row.names(thresholds) <- (c("min","max"))
```


```{r}
# map the forecast summaries one at a time 
forecast.maps <- MapForecast(forecasts.n, base.map, thresholds)

# to access one of the individual forecast maps: 
forecast.maps$`3-day`
```

