---
title: "CLIMARK Charts Tutorial: Plotting aWhere weather time series data for an X,Y location"
author: "Victoria Scholl"
date: "6/19/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      results = 'hide')
```

# Introduction

This document describes how to generate aWhere weather data charts for a given location and timespan. Current/forecasted weather data is pulled from the aWhere API along with long-term normals for that same time span. Displaying line plots for variables of interest such as maximum temperature and P/PET allows us to observe weather trends and how current conditions compare to the past. 

# Code

## Install / load R packages

First, install the aWhere R packages. 

```{r install_aWhere_packages, eval=FALSE}
# install aWhere R packages
library(devtools)
devtools::install_github("aWhereAPI/aWhere-R-Library")
devtools::install_github("aWhereAPI/aWhere-R-ChartLibrary")
```

Load the R packages that contain functions used in this script. 

```{r load_packages}
# load required packages 
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(aWhereAPI)
library(zoo)
library(aWhereCharts)
library(curl)
```

## Define input files and parameters

Define your working directory. This is the place on your computer where your input files are located for this tutorial, and it is also where the output files will be written. Instead of typing out the entire path to every file that we need ("absolute" paths), you can just use their filenames or locations within this working directory ("relative" paths) since we are setting this working directory as our starting point for the rest of the tutorial. 

```{r set.wd}
# define your working directory - where input files are located and outputs will be saved
working.dir <- "~/Documents/aWhere/"

# set the working directory
setwd(working.dir)
```

Use the *source* function to load R functions that are located on your local machine. For this tutorial, there are a few functions within the *supporting_functions.R* file. 

```{r load.functions}
# load external R functions in local file
source("supporting_functions.R")
```

To pull data from the aWhere API, you'll need a Consumer Key and Secret, like a username and password. Put these in a text file with the following format: line 1 is the Consumer Key, line 2 is the Consumer Secret, and line 3 is a blank line. Set the *credentials.file* variable equal to the name of this text file. Place your credentials file in your working directory. 

```{r set.credentials.file}
# filename containing your aWhere credientials (key and secret)
credentials.file <- "credentials.txt"
```

What location do you want to produce weather data charts for? Specify the latitude and longitude, along with a name describing the place. This name will be used for the chart names and output file names.

```{r set.location}
# latitude, longitude, and name of location 
lat <- 3.2
lon <- 37.0
location.name <- paste("Shamba1") 
```

What timespan to you want to acquire weather data for? Specify the years and dates here. *Years* is a vector with the starting year in position 1 and the ending year in position 2, for the long-term analysis. 

The starting and ending days should have the format "YYYY-MM-DD" and should not exceed a period greater than 365 days. If you want the ending date to be *n* days from now (including some forecasted weather data), add or subtract *n* to the current date (which can be acquired using the *Sys.Date()* function). In the code below, *n* = 7 so the end date is seven days from now. 

```{r set.years.days}

# starting and ending years and days for the time series. 
# "years"" is a vector with the starting year in position 1 and the ending
# year in position 2, for the long-term analysis. 
years <- c(2010, 2017)

#starting and ending days with format "YYYY-MM-DD"
day.start <- "2018-03-01"
day.end <- "2018-03-08" # specific date
day.end <- as.character(Sys.Date() + 7) # today plus n

# combine the days into a single vector 
days <- c(day.start, day.end)
```

Some of the charts will require additional parameters. For instance, effective precipitation can be calculated using a threshold (*eP* in the code below). Rolling average calculations require a window size or number of days to average over (*roll.avg* in the code below). 

```{r extra.parameters}
# effective precip amount for consistency across plots
eP <- 30    

# size of rolling average window for consistency across plots
roll.avg <- 30
```


## Processing steps 

The duration for our time range cannot exceed 364 days. Let's make sure our day range is an acceptable length. 

```{r check.duration}
# Check if time range is more than 365 days
duration <- as.Date(days[2]) - as.Date(days[1])

if(duration > 364) { 
  print ("Period too long")
}
```

Pull the aWhere weather data for the specified location and timespan. 

The R function that we use to pull the data is *generateaWhereDataset*, which comes from the *aWhereCharts* R package. It's good practice to specify which package each function is from, since multiple packages may have different functions with the same names. The syntax for doing this is to use two colons between the package and function names, like this: **package::function()**. 

```{r pull.weather.data}
# load the aWhere API credentials file to authenticate yourself
aWhereAPI::load_credentials(paste0(working.dir,credentials.file)) 

# pull the datasets: query the aWhere API
weather.df <- aWhereCharts::generateaWhereDataset(lat = lat, lon = lon, 
                                                  day_start = days[1], 
                                                  day_end = days[2], 
                                                  year_start = years[1], 
                                                  year_end = years[2])
```
Let's take a look at the weather data using the *head* function from the *utils* R package. 

```{r show.weather.data, results=TRUE}
# select the first five columns of the data frame using [,1:5]
# and show the first 10 rows using the "n" argument
utils::head(weather.df[,1:5], n = 10)
```

Write the weather data to a .csv file for further analysis. Use the latitude, longitude, date range, and year range in the filename so future-you will know exactly what's inside this file. 

```{r write.weather.data}
# write data to .csv file 
utils::write.csv(weather.df, 
                 file = paste(location.name, 
                              base::paste(lat, lon, sep="_"),
                              base::paste(days, collapse="_"),
                              base::paste(years, collapse="_"),
                              ".csv", sep="_"))
```

# Charts

Within the *aWhereCharts* package, there are two functions used here to create
time series charts. **generateaWhereChart** makes line plots comparing the current and long-term normals for each variable. **generateaWhereStdDevChart** plots aWhere weather data with one standard 
deviation's shading above and below the long-term normalline. This function is meant to help users understand whether the current weather conditions are significantly outside the norm. 

Plots with standard deviation are named with "1" at the end of the variable, and those without standard deviation are named with "2". 

## Maximum temperature with standard deviation

Contruct the title for this chart using the variable name, location name, and the latitude/longitude.

```{r max.temp.1}
# construct title
max.temp.1.title <- paste0("Maximum Temp w StdDev ", location.name,
                           " (", lat, ", ", lon, ")")
```

Generate the plot using the *generateaWhereStdDevChart* function from the *aWhereCharts* package. 

```{r}
# generate the plot 
max.temp.1 <- aWhereCharts::generateaWhereStdDevChart(data = weather.df, 
                                                      variable = "maxTemp", 
                                                      title = max.temp.1.title)
```

Display the plot and write it to .JPEG file using the WriteJpeg function, an external R function in the "supporting_functions.R" file.

The "current" observations (blue line) are within the *day.start* and *day.end* range that you set earlier. The "LTN" (long-term normal) observations (orange line) are calculated over the course of the year range that you set earlier. The orange shading around the orange line is one standard deviation of the observations to illustrate variability. If your day range extends into the future (beyond today), a black dashed line indicates the start of the forecasted data. 

```{r results=TRUE}
# display plot
max.temp.1 
```

```{r}
# write .jpeg
WriteJpeg(plt = max.temp.1, plt.title = max.temp.1.title)
```

In the following sections, these same steps are performed for a series of other variables in the aWhere weather data set. 

## Minimum temperature with standard deviation 

```{r min.temp.1}
min.temp.1.title <- paste0("Minimum Temp w StdDev ", location.name,
                           " (", lat, ", ", lon, ")")

min.temp.1 <- aWhereCharts::generateaWhereStdDevChart(weather.df, 
                                                      "minTemp", 
                                                      min.temp.1.title)
# display plot
min.temp.1 

# write to file
WriteJpeg(plt = min.temp.1, plt.title = min.temp.1.title)
```

## Potential evapotranspiration (PET) with standard deviation

```{r pet.2}
pet.1.title <- paste0("PET w StdDev ", location.name,
                      " (", lat, ", ", lon, ")")

pet.1 <- aWhereCharts::generateaWhereStdDevChart(weather.df, 
                                                 "pet", 
                                                 pet.1.title)

# display plot
pet.1 

# write to file
WriteJpeg(plt = pet.1, plt.title = pet.1.title)
```

## Daily precipitation with standard deviation

```{r}
precip.1.title <- paste0("Daily Precipitation w StdDev ", location.name,
                         " (", lat, ", ", lon, ")")

precip.1 <- aWhereCharts::generateaWhereStdDevChart(weather.df, 
                                                    "precipitation",  
                                                    title = precip.1.title)
# display plot
precip.1 

# write to file
WriteJpeg(plt = precip.1 , plt.title = precip.1.title)
```

## Daily precipitation without standard deviation  

```{r}
precip.2.title <- paste0("Daily Precipitation ", location.name,
                         " (", lat, ", ", lon, ")")

precip.2 <- aWhereCharts::generateaWhereChart(weather.df,
                                              "precipitation",  
                                              title = precip.2.title)

# display plot
precip.2 

# write to file
WriteJpeg(plt = precip.2, plt.title = precip.2.title)
```

## Accumulated Precipitation with StdDev but no Effective Precipitation 

```{r}
no.eprecip.1.title <- paste0("Accumulated Precipitation w StdDev ", location.name,
                             " (", lat, ", ", lon, ")")
no.eprecip.1 <- aWhereCharts::generateaWhereStdDevChart(weather.df, 
                                                        "accumulatedPrecipitation", 
                                                        no.eprecip.1.title)
# display plot
no.eprecip.1

# write to file
WriteJpeg(plt = no.eprecip.1, plt.title = no.eprecip.1.title)
```


## Precipitation and Effective Precipitation, Accumulated 

```{r}
eprecip.1.title <- paste0("Precipitation and Effective Precipitation, 
                          Accumulated w Std Dev", location.name, " (", 
                          lat, ", ", lon, ")")

eprecip.1 <- aWhereCharts::generateaWhereStdDevChart(weather.df, 
                                                     "accumulatedPrecipitation", 
                                                     eprecip.1.title,
                                                     e_precip = TRUE, 
                                                     e_threshold = eP)
# display plot
eprecip.1

# write to file
WriteJpeg(plt = eprecip.1, plt.title = eprecip.1.title)
```

## Accumulated Precipitation 

```{r}
acc.precip.2.title <- paste0("Accumulated Precipitation ", location.name, " (", 
                             lat, ", ", lon, ")")

acc.precip.2 <- aWhereCharts::generateaWhereChart(weather.df, 
                                                  "accumulatedPrecipitation", 
                                                  acc.precip.2.title)
# display plot
acc.precip.2

# write to file
WriteJpeg(plt = acc.precip.2, plt.title = acc.precip.2.title)
```

## Accumulated PET with standard deviation
```{r}
acc.pet.1.title <- paste0("Accumulated PET w StdDev ", location.name, " (", 
                          lat, ", ", lon, ")")

acc.pet.1 <- aWhereCharts::generateaWhereStdDevChart(weather.df, 
                                                     "accumulatedPet", 
                                                     acc.pet.1.title)  
# display plot
acc.pet.1

# write to file
WriteJpeg(plt = acc.pet.1, plt.title = acc.pet.1.title)
```


##  P/PET 
(P/PET rarely is interpretable on a daily chart)

```{r}
ppet.2.title <- paste0("P PET ", location.name, " (", 
                       lat, ", ", lon, ")")

ppet.2 <- aWhereCharts::generateaWhereChart(weather.df, 
                                            "ppet", 
                                            ppet.2.title)
# display plot
ppet.2

# write to file
WriteJpeg(plt = ppet.2, plt.title = ppet.2.title)
```


## 30 day rolling average eP/PET and P/PET 

No eprecip/PET shows up if all rainfall events are less than the effective precipitation threshold. 

```{r}
rolling.avg.ppet.2.title <- "30 day rolling avg eP PET and P PET "

rolling.avg.ppet.2 <- aWhereCharts::generateaWhereChart(weather.df, 
                                                        "rollingavgppet",
                                                        rolling.avg.ppet.2.title,
                                                        e_precip = TRUE, 
                                                        e_threshold = eP, 
                                                        rolling_window = roll.avg)
# display plot
rolling.avg.ppet.2

# write to file
WriteJpeg(plt = rolling.avg.ppet.2, plt.title = rolling.avg.ppet.2.title)
```

## Multiplot 

Select any of the above charts to display together in a multiplot using their variable names: 
  - max.temp.1
  - min.temp.1
  - pet.1
  - precip.1
  - precip.2
  - no.eprecip.1
  - eprecip.1
  - acc.precip.2
  - acc.pet.1
  - ppet.2
  - rolling.avg.ppet.2
  
```{r multiplot}
# set the graphics device parameters to write a .JPEG
jpeg(paste0(location.name,"_4chart.jpeg"), 
     width = 12, height = 6, 
     units = 'in', res = 500)

# generate the multiplot & write to JPEG
aWhereCharts::generateMultiplot(eprecip.1, rolling.avg.ppet.2, max.temp.1, pet.1, 
                  cols = 2, fontsize = 15, 
                  title = paste0("Current vs LTN at ", 
                                 location.name," (", lat, ", ", lon, ")", 
                                 "   eP = ",eP,"mm"))

# close the current plot object
dev.off()
```

  
  

