---
title: 'CLIMARK Tutorial: Statistics and Histograms'
author: "Victoria Scholl"
date: "6/19/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      results = 'hide')
```

# Introduction

# Code

## Install / load R packages

First, install the aWhere R packages. 

```{r install_aWhere_packages, eval=FALSE}
# install aWhere R packages
library(devtools)
devtools::install_github("aWhereAPI/aWhere-R-Library")
devtools::install_github("aWhereAPI/aWhere-R-ChartLibrary")
```

Load the R packages that contain functions used in this script. 

```{r load_packages}
library(dplyr)
library(tidyverse) 
library(tibble)
library(tidyr)
library(ggplot2)
library(ggmap)
library(ggthemes)
library(wicket) 
library(aWhereAPI)
library(aWhereCharts)
```

## Define input files and parameters

Define your working directory. This is the place on your computer where your input files are located for this tutorial, and it is also where the output files will be written. Instead of typing out the entire path to every file that we need ("absolute" paths), you can just use their filenames or locations within this working directory ("relative" paths) since we are setting this working directory as our starting point for the rest of the tutorial. 

```{r set.wd}
# define your working directory - where input files are located and outputs will be saved
working.dir <- "~/Documents/aWhere/"

# set the working directory
setwd(working.dir)
```

Use the *source* function to load R functions that are located on your local machine. For this tutorial, there are a few functions within the *supporting_functions.R* file. 

```{r load.functions}
# load external R functions in local file
source("supporting_functions.R")
```

Specify the weather data directory and file name. 

```{r}
# specify the weather data directory and file name
weather.dir <- "climark_work_csvs/" 
weather.name <- "180609_past30.csv"
```

Specify the template data filename containing the geographic location data.

```{r}
# define template data filename
template.file <- "CLIMARKonlyWardTemplate.csv"
```

If you want to write the histogram plots to image files, set *write.hist* to TRUE. 

```{r}
# write histograms to image files 
write.hist = TRUE
```

To select subarea(s) of interest, list their names in this vector. For now, these subareas are limited to ward names. To generate a forecast for the entire region instead, set *subarea.select* to "ENTIRE_REGION". 

```{r}
subarea.select <- "KARARE" 
subarea.select <- c("KARARE", "GOLBO")
subarea.select <- "ENTIRE_REGION"
```

Define the bin ranges and increments to create tabular summaries of histogram data.

```{r}
# precipitation
bins.precip <- c(seq(from = 0, to = 300, by = 5), Inf) 

# P/PET
bins.ppet <- c(0, 0.4, 0.6, 0.7, 0.8, 0.9, 1.0, 1.1, 1.2, 1.4, 1.6, 2.0, Inf) 
```

When plotting the weather data, we will first display a geographic base map. Define the latitude, longitude, and zoom values for the base map. For the entire CLIMARK region, a latitude of 2.5, longitude of 38, and zoom of 7 are appropriate values. 

```{r}
# base map location and zoom values for mapping the forecast data. 
map.lat <- 2.5
map.lon <- 38
map.zoom <- 7
```


## Processing steps

Read the weather data.

```{r}
# combine the directory and file name
weather.file <- paste(working.dir, weather.dir, weather.name, sep="")

# read the weather data 
weather.df <- read.csv(weather.file)
```

Read the template data. Filter the data to keep only the grid cells within the subarea of interest. Merge the weather and template data by their location ID's into a single data frame. 

```{r}
# read the template data. remove columns that are not necessary.
template.df <- read.csv(paste0(working.dir,template.file)) %>% 
  dplyr::select( -c(shapewkt, longitude, latitude ))

# filter weather data for only the grid locations within the template data 
get.wards.area <- weather.df %>% 
  dplyr::filter(locationid %in% template.df$locationid)

# merge the weather data with and template data (wards/constituen)
weather.template.df <- merge(get.wards.area, 
                          template.df, by = "locationid")
```

Let's take a look at the combined data set. Both the weather data and template data are available for each grid cell or location ID. 

```{r results=TRUE}
head(weather.template.df %>% 
       dplyr::select(locationid, latitude, longitude, CSUMPRE, 
                     CPOVRPR, WARD_NAME))
```

Write the combined data set to a .csv file. 

```{r}
# construct output filename for weather + template data
weather.template.df.file <- paste("weather+template",
                                  weather.name,
                                  sep = "_")

# write the combined weather and template data to .csv file
write.csv(weather.template.df, 
          file = weather.template.df.file)
```

Filter the data set for subarea(s) of interest and write this clipped data 
set to file. It can become a template for pulling data from the aWhere API.

```{r}
if (!identical(subarea.select, "ENTIRE_REGION")){ 
  
  weather.template.df <- weather.template.df %>% 
    dplyr::filter(WARD_NAME %in% subarea.select) 
  
  write.csv(weather.template.df, file = paste("weather+template_clip_",
                                              weather.name,
                                              sep = "_"))
} 
```

Let's take a look at the combined data set after the subarea filtering. If we are using the "ENTIRE_REGION", it will look the same as before. If we are restricting the data set to a ward or multiple wards, the "WARD_NAME" column will have those restricted values.

```{r results=TRUE}
head(weather.template.df %>% 
       dplyr::select(locationid, latitude, longitude, CSUMPRE, 
                     CPOVRPR, WARD_NAME))
```

## Calculate statistics 

For each ward in the data set, let's calculate some statistics such as the mean, max, and standard deviation for selected weather variables. We will calculate these same statistics across the entire CLIMARK region as well, and join these results together for comparison. The number of grid cells per subarea is calculated as well. 

```{r}
# calculate stats across subareas
subarea.stats <- weather.template.df %>%
  dplyr::group_by(WARD_NAME) %>% 
  dplyr::summarise(avg_CSUMPRE = mean(CSUMPRE),
                   max_CSUMPRE = max(CSUMPRE),
                   sd_CSUMPRE = sd(CSUMPRE),
                   avg_LTNsumPre = mean(LTNSUMP),
                   max_LTNsumPre = max(LTNSUMP),
                   sd_LTNsumPre = sd(LTNSUMP),
                   avg_D_CLTNSUMPRE = mean(DFLTSUM),
                   max_D_CLTNSUMPRE = max(DFLTSUM),
                   sd_D_CLTNSUMPRE = sd(DFLTSUM),
                   avg_CP_PET = mean(CPOVRPR),
                   max_CP_PET = max(CPOVRPR),
                   sd_CP_PET = sd(CPOVRPR),
                   avg_LTNP_PET = mean(LTNASPO),
                   max_LTNP_PET = max(LTNASPO),
                   sd_LTNPPET = sd(LTNASPO),
                   avg_D_CLTNP_PET = mean(DFLTPVP),
                   max_D_CLTNP_PET = max(DFLTPVP),
                   sd_D_CLTNP_PET = sd(DFLTPVP),
                   avg_CAvgMinT = mean(CAvgMinT),
                   max_CAvgMinT = max(CAvgMinT),
                   sd_CAvgMinT = sd(CAvgMinT),
                   avg_CAvgMaxT = mean(CAvgMaxT),
                   max_CAvgMaxT = max(CAvgMaxT),
                   sd_CAvgMaxT = sd(CAvgMaxT),
                   n_grids = n())

# calculate the stats across the entire region as a single entry in the table
# this serves as a summary across the entire region
region.stats <- weather.template.df %>%
  dplyr::summarise(avg_CSUMPRE = mean(CSUMPRE),
                   max_CSUMPRE = max(CSUMPRE),
                   sd_CSUMPRE = sd(CSUMPRE),
                   avg_LTNsumPre = mean(LTNSUMP),
                   max_LTNsumPre = max(LTNSUMP),
                   sd_LTNsumPre = sd(LTNSUMP),
                   avg_D_CLTNSUMPRE = mean(DFLTSUM),
                   max_D_CLTNSUMPRE = max(DFLTSUM),
                   sd_D_CLTNSUMPRE = sd(DFLTSUM),
                   avg_CP_PET = mean(CPOVRPR),
                   max_CP_PET = max(CPOVRPR),
                   sd_CP_PET = sd(CPOVRPR),
                   avg_LTNP_PET = mean(LTNASPO),
                   max_LTNP_PET = max(LTNASPO),
                   sd_LTNPPET = sd(LTNASPO),
                   avg_D_CLTNP_PET = mean(DFLTPVP),
                   max_D_CLTNP_PET = max(DFLTPVP),
                   sd_D_CLTNP_PET = sd(DFLTPVP),
                   avg_CAvgMinT = mean(CAvgMinT),
                   max_CAvgMinT = max(CAvgMinT),
                   sd_CAvgMinT = sd(CAvgMinT),
                   avg_CAvgMaxT = mean(CAvgMaxT),
                   max_CAvgMaxT = max(CAvgMaxT),
                   sd_CAvgMaxT = sd(CAvgMaxT),
                   n_grids = n()) %>% 
      dplyr::mutate(WARD_NAME = "ENTIRE REGION") %>%
      dplyr::select(WARD_NAME, n_grids, everything())

# combine the ward-specific stats with the overall region calculation
stats.out <- rbind(region.stats,
                   subarea.stats)
```

Let's take a look at the statistics data.

```{r results=TRUE}
head(stats.out[,1:5])
```

Write these statistics to a .csv file.

```{r}
write.csv(stats.out,
          paste("stats_by_subarea",
                weather.name,
                sep="_"))
```


# Histograms 

Visualize the data using the **generateaWhereHistogram** function within the **aWhereCharts** package. Compare the current versus long-term observations of each variable. 

## Precipitation 

```{r}
hist.title <- paste("Histogram Precipitation", 
                    tools::file_path_sans_ext(weather.name), sep = " ")

aWhereCharts::generateaWhereHistogram(data = weather.template.df, 
                        variable = "CSUMPRE", 
                        title = hist.title, 
                        xlabel = "mm", 
                        compare = TRUE, 
                        compare_var = "LTNSUMP")

# write histogram to file 
if (write.hist == TRUE) {
  ggplot2::ggsave(paste0(filename = hist.title, ".png"),
       device = "png")
}
```

## Minimum temperature

```{r}
hist.title <- paste("Histogram Min Temp", 
                    tools::file_path_sans_ext(weather.name), sep = " ")

generateaWhereHistogram(data = weather.template.df, 
                        variable = "CAvgMinT", 
                        title = hist.title, 
                        xlabel = "Deg C", 
                        compare = TRUE, 
                        compare_var = "LTAvgMnT")

# write histogram to file 
if (write.hist == TRUE) {
  ggplot2::ggsave(paste0(filename = hist.title, ".png"),
         device = "png")
}
```

## Maximum temperature

```{r}
hist.title <- paste("Histogram Max Temp", 
                    tools::file_path_sans_ext(weather.name), sep = " ")

generateaWhereHistogram(data = weather.template.df, 
                        variable = "CAvgMaxT", 
                        title = hist.title, 
                        xlabel = "Deg C", 
                        compare = TRUE, 
                        compare_var = "LTAvgMxT")

# write histogram to file 
if (write.hist == TRUE) {
  ggplot2::ggsave(paste0(filename = hist.title, ".png"),
         device = "png")
}
```

## P / PET

Sometimes, the weather observations have a very large range with very few extreme values. This can result in a histogram with a large tail, which can make it difficult to interpret the region containing most of the data. To avoid this, we will clip the extreme values of the current P/PET and long-term average P/PET to a range of [0,2] using the ClipValues function, found in the "supporting_functions.R" file. 

```{r}
# clip the extreme values of CPOVRPR (current P/PET) and LTNASPO 
# long-term average P/PET and place these clipped values in new columns,
# "ctemp" and "LTNtemp"
weather.template.df$ctemp <- ClipValues(weather.template.df$CPOVRPR, 
                                        max.thresh = 2)
weather.template.df$LTNtemp <- ClipValues(weather.template.df$LTNASPO, 
                                        max.thresh = 2)

# use this separate histogram function for now to plot P/PET
source(paste0(working.dir, "function_generateaWhereHistogramPET.R"))

# construct a descriptive title
hist.title <- paste("Histogram PPET", 
                    tools::file_path_sans_ext(weather.name), sep = " ")
  
generateaWhereHistogramPET(data = weather.template.df, 
                           "ctemp", 
                           title = hist.title, 
                           xlabel = "P/PET", 
                           compare = TRUE, 
                           compare_var = "LTNtemp")

# write histogram to file 
if (write.hist == TRUE) {
  ggplot2::ggsave(paste0(filename = hist.title, ".png"),
         device = "png")
}
```

# Tabular summaries of histogram data

In addition to visualizing the distribution of weather variables using histogram plots, it is helpful to summarize the number of observations within the bins using a table. 

Let's count how many grid cells belong to each precipitation level. A list of precipitation bins was defined earlier in the script (*bins.precip*) from 0 mm to 300 mm. 

```{r results=TRUE}
# take a look at the bins
bins.precip
```

If our actual precipitation data does not actually stretch up to 300 mm, we can remove some of those bins.

```{r}
# if the initial precipitation bins extend beyond the actual range of precip
# data, remove the extra bins
bins.precip <- bins.precip[bins.precip < (max(weather.template.df$CSUMPRE) + 5)] 
```

```{r results=TRUE}
# take a look at the bins after removing the ones that will be empty
bins.precip
```

Let's create empty columns in our data frame to keep track of the precipitation bin (*bin.precip*) and the extent of the bin, "0-5 mm", "5-10 mm", etc. (*bin.range.precip*). 

```{r}
# add a column for which bin/ bin range each grid falls into
weather.template.df$bin.precip <- NA
weather.template.df$bin.range.precip <- NA
```

For every grid cell, keep track of which precipitation bin and bin range it belongs to. For instance, a precipitation value of 7.6mm will fall into bin number 2, which has a range of 5-10 mm. 

```{r}
# loop through each bin and populate the appropriate values
for(b in 1:(length(bins.precip)-1)){
  
  # indices of entries that fall in the current bin
  idx <- weather.template.df$CSUMPRE >= bins.precip[b] & 
                   weather.template.df$CSUMPRE < bins.precip[b+1]
  
  # add the bin number to each row
  weather.template.df$bin.precip[idx] <- b
  
  # add the bin range to each row 
  weather.template.df$bin.range.precip[idx] <- paste(as.character(bins.precip[b]),
                                      " - ",
                                      as.character(bins.precip[b+1]),
                                      sep="")
}
```

Let's count the total number of grids per precipitation bin, in addition to the percentage of total grids. By regrouping our data, we can display a nice tabular summary that corresponds to what the precipitation histogram showed. 

```{r results=TRUE}
# add columns for number of grids per precip level
weather.template.df <- weather.template.df %>% 
  dplyr::group_by(bin.precip) %>%
  dplyr::mutate(grid.count.precip = n(),
         grid.percent.precip = 100 * n() / nrow(weather.template.df))

# nest the data within each bin and show a summary
weather.template.nested.precip <- weather.template.df %>% 
  dplyr::group_by(bin.precip, 
           bin.range.precip,
           grid.count.precip, 
           grid.percent.precip) %>%
  tidyr::nest() %>% 
  dplyr::arrange(bin.precip)

# take a look at the nested data. 
# this is essentially the histogram in tabular form.
# get the nested data out of the way,
grids.per.precip.level <- weather.template.nested.precip %>% 
                                                 dplyr::select(-data)

# and use the "head" function to take a look at the first 6 rows
head(grids.per.precip.level)
```

Write the grid count per precip level summary to a .csv file. 

```{r}
# write precip number of grids per precip level to .csv
write.csv(grids.per.precip.level,
          "grids_per_precip_level.csv")
```


We can perform the same steps to generate a tabular summary for the P/PET variable: 

```{r results=TRUE}
# take a look at the P/PET bins
bins.ppet
```

```{r}
# add a column for which bin the grid falls into
weather.template.df$bin.ppet <- NA
weather.template.df$bin.range.ppet <- NA

# loop through each bin and populate the appropriate values
for(b in 1:(length(bins.ppet)-1)){
  
  # indices of entries that fall in the current bin
  idx <- weather.template.df$CPOVRPR >= bins.ppet[b] & 
    weather.template.df$CPOVRPR < bins.ppet[b+1]           
  
  # add the bin number to each row
  weather.template.df$bin.ppet[idx] <- b
  
  # add the bin range to each row 
  weather.template.df$bin.range.ppet[idx] <- paste(as.character(bins.ppet[b]),
                                           " - ",
                                           as.character(bins.ppet[b+1]),
                                           sep="")
}

# add columns for number of grids per ppet level
weather.template.df <- weather.template.df %>% 
  dplyr::group_by(bin.ppet) %>%
  dplyr::mutate(grid.count.ppet = n(),
         grid.percent.ppet = 100 * n() / nrow(weather.template.df))

weather.template.nested.ppet <- weather.template.df %>% 
  dplyr::group_by(bin.ppet, 
           bin.range.ppet,
           grid.count.ppet, 
           grid.percent.ppet) %>%
  tidyr::nest() %>% 
  dplyr::arrange(bin.ppet)
```

```{r results=TRUE}
# take a look at the nested data
# this is essentially the P/PET histogram in tabular form.
# get the nested data out of the way,
grids.per.ppet.level <- weather.template.nested.ppet %>% 
                           dplyr::select(-data)

# and use the "head" function to take a look at the first 6 rows
head(grids.per.ppet.level)
```

```{r}
# write number of grids per ppet level to .csv
write.csv(grids.per.ppet.level,
          "grids_per_ppet_level.csv")
```


# Ward narrative 

```{r}
# To create a narrative about the percentage of a ward receiving 
# a given precip or p/pet level, we can subset the larger data frame 
# and see the distribution of grids cells at different precipitation / ppet levels
ward.select = "TURBI"
ward.select = "KARARE"

# filter the data for the ward of interest
ward.df <- weather.template.df %>% 
  dplyr::filter(WARD_NAME %in% ward.select) 

# count total number of grids in ward 
ward.grid.count <- nrow(ward.df)

# calculate the number of grids and % of grids per precip bin
ward.df <- ward.df %>%
  dplyr::group_by(bin.precip) %>% 
  dplyr::mutate(bin.grid.count.ward = n(),
                bin.grid.percent.ward = 100 * (bin.grid.count.ward / 
                                                 ward.grid.count)) %>%
  dplyr::group_by(bin.precip,
                  bin.range.precip,
                  bin.grid.count.ward,
                  bin.grid.percent.ward) %>%
  tidyr::nest() %>% 
    dplyr::arrange(bin.precip)
```

```{r results=TRUE}
# take a look at the grid count and percentages per precip level
head(ward.df %>% dplyr::select(-data))
```

# Mapping the weather variables 

```{r}
# create the base map using the parameters defined earlier
base.map = ggmap::get_map(location = c(lon = map.lon, 
                                lat = map.lat), 
                   zoom = map.zoom, 
                   color = "bw")

# display map of region
gg.map <- ggmap(base.map)
#gg.map


# clip the extreme values of selected variables to map
weather.template.df$cPovPET <- ClipValues(weather.template.df$CPOVRPR, 
                                       max.thresh = 2)
weather.template.df$cLTNPPET <- ClipValues(weather.template.df$LTNASPO, 
                                          max.thresh = 2)
weather.template.df$aPre <- ClipValues(weather.template.df$CSUMPRE, 
                                           max.thresh = 300)
weather.template.df$aLTNPRE <- ClipValues(weather.template.df$LTNSUMP, 
                                       max.thresh = 400)
weather.template.df$aDinPre <- ClipValues(weather.template.df$DFLTSUM, 
                                          max.thresh = 250,
                                          min.thresh = -250)

ggmap.df <- weather.template.df

# Expand wkt to format usable by ggplot
polygon.df = as.tibble(wicket::wkt_coords(ggmap.df$shapewkt))
polygon.df$aPre <- ggmap.df$aPre[polygon.df$object]
polygon.df$cPovPET = ggmap.df$cPovPET[polygon.df$object]
polygon.df$aDinPre = ggmap.df$aDinPre[polygon.df$object]

# call the MakeMap function from the "supporting_functions.R" script to create a map
# for each of the specified variables
```

```{r results=TRUE}
# Precipitation map
MakeMap(df = polygon.df, v = "pre", base.map, 
        tools::file_path_sans_ext(weather.name), write.file = TRUE)
```

```{r results=TRUE}
# P/PET map
MakeMap(df = polygon.df, v = "ppet", base.map, 
        tools::file_path_sans_ext(weather.name), write.file = TRUE)
```

```{r results=TRUE}
# P LTN map
MakeMap(df = polygon.df, v = "pltn", base.map, 
        tools::file_path_sans_ext(weather.name), write.file = TRUE)
```

